Variety in the Implementation of Nonlinear Least Squares Program Codes

### 2022-01-20 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
more changes to Variety

Given a function

@@

- inequality constraints, which in general may be written as 
  $$ C_k(x) >=0 $$, which can be written as a vector form $$ C(x) >= 0 $$
  A special and important case is that of **linear inequalities** $$ A x >= b $$,
  that is, $$ C(x) = A x - b $$.

@@

An overview of the method is provided by Wikipedia:

\url{https://en.wikipedia.org/wiki/Gauss%E2%80%93Newton_algorithm}

The method was described by C.F.Gauss in 1809 in *Theoria motus corporum 
coelestium in sectionibus conicis solem ambientum*.

In calculus we learn that a stationary point (local maximum, minimum
or saddle point) of a function $f$ occurs where its gradient (or first
derivative) is zero. In multiple dimensions, this is the same as
having the gradient $g$ stationary. The so-called Newton method
uses the second derivatives in the Hessian matrix $H$ defined as

@@

and we iterate until there is no change in $x$.

There are some adjustments when our function $f(x)$ can be written as a sum of squares

The final vector product form is the one I favour because it is the least fussy to write and read.
In particular, when we try to evaluate the Hessian of this sum of squares function, we see that it is

@@

then the gradient is

@@

Arguing that the residuals should be "small", Gauss proposed that the 
Newton equations could be approximated by ignoring
the second term (using elements of the residual times second derivatives 
of the residual).  This gives the **Gauss-Newton**
equations (cancelling the factor 2)

@@

@@

with respect to the (scalar) parameter $\alpha$. That one-dimensional minimization can
take quite a lot of computational effort, however, and typically a value of $\alpha$ that
permits a suitable reduction in the sum of squares $f(x)$ is accepted, and the iteration repeated
from

@@

- **A** Compute $J$ and $g$, set up Marquardt equations and find $\delta$

- **B** compute new set of parameters $$x_{new} = x + \delta $$

@@

Many practitioners dislike using IF statements (comparisons) of floating point
numbers. Tolerances are often suggested to define "small", but this invites users
to make decisions that lead to poor outcomes and which ignore the precision and
other characteristics of the arithmetic at hand.
There are, of course, well-written codes that address the issues of what "small" means,
but such code is generally quite involved.
I have never seen a case where the $OFFSET$ approach caused difficulties, and it
is trivial to code. When $x$ is tiny, we compare $OFFSET$ to 
itself and declare equality. When $x$ is very large, $OFFSET$ does not change it and
we compare $x$ with a value that is identical to it.

@@

<!-- ?? Check the statements about how things are done and redo tests -->
Note that we form $J^t J$ in `nlsr::nlfb()` to provide the $D$ matrix for 
the scaled stabilization, 
which is quite a lot of computational work. The same program, however, 
uses a QR approach to solving the linearized least squares sub-problem.
To avoid the formation of $D$, an experimental package
`nlsralt` was been built to add functions `nlxbx()` and `nlfbx()` that
use ONLY the fixed element (i.e., unit matrix) stabilization. A very
preliminary timing ([2021-6-26]) using the Hobbs unscaled problem shows
a slight improvement for the simplified version in timing with equivalent
solution. ?? check

@@

There have been many proposed approaches to nonlinear least squares. The following
is a non-exhaustive sample.

@@

@Jones70Spiral seems only to be mentioned in a single paper. There is a flowchart
given but I have never been able to find a code (supposedly Fortran V for the Univac
1108.) The motivations are to avoid both matrix equation solutions and one-dimensional
line-search minimization. Jones mentions matrix inversion, and it is not clear whether
he truly forms the inverse. The main idea in the paper (and presumably the code) is
to use a geometric spiral that mimics the Marquadt search trajectory as $\lambda$
is increased. It would be interesting to actually try some of the ideas, which do,
as far as I can determine, have a reasonable basis.

@@

Newton (and his sometime associated worker Raphson) used the algorithm we attribute
to them applied to a very special case and in a manner unrecognizable as that described
above.  (See \url{https://en.wikipedia.org/wiki/Newton%27s_method}).

The main attraction of the Gauss-Newton method is that the Jacobian only needs first
derivatives, which are generally much, much easier to compute than the full second
derivatives of the Hessian. However, there have been suggestioned
Newton-like approaches, possibly for special types of models where some short cut
to the Hessian is available. These may or may not make sense in the context of
a given user or group of users. I regard these approaches as part of the
computational infrastructure of the special cases rather than as general nonlinear
least squares methods.

?? MaxLik?? others -- Charles Geyer

https://www.stat.umn.edu/geyer/trust/

CRAN trust package??

?? had something for exponential models??

@@

The central idea of Hartley's method is a line search and that of Marquardt a
parametrized stabilization that can be shown to reduce a sum of squares for large
enough stabilization parameter. There is nothing to prevent use of a line
search additional to the Marquardt stabilization, and there are, of course, 
several reasonable choices for line search. Also we could solve the (possibly
stabilized) Gauss-Newton equations by several quite different numerical linear
algebra techniques, or even approximations. In certain situations, some of these
choices may be important, but here I will not pursue them further.

@@

The sources of variety in implementation include:

@@

- sequential or full creation of the Jacobian and residual, since
  it may be done in parts

>>>>

Given a function that can be written as a sum of squared residuals

@@

- inequality constraints, which in general may be written as 
  $$ C_k(x) >=0 $$ which can be written as a vector form $$ C(x) >= 0 $$
  A special and important case is that of **linear inequalities** $$ A x >= b $$
  that is, $$ C(x) = A x - b $$

@@

An overview of this method is provided by Wikipedia: 
\url{https://en.wikipedia.org/wiki/Gauss%E2%80%93Newton_algorithm}
The method was described by C.F.Gauss in 1809 in *Theoria motus corporum 
coelestium in sectionibus conicis solem ambientum*.

In calculus we learn that a stationary point (local maximum, minimum
or saddle point) of a function $f$ occurs where its gradient (or first
derivative) is zero. In multiple dimensions, this is the same as
having the gradient $g$ stationary. The so-called **Newton method**
uses the second derivatives in the Hessian matrix $H$ defined as

@@

and we iterate until there is no change in $x$, assuming that the method does indeed
converge for the problem at hand. Newton (and his sometimes associated worker Raphson) 
used the algorithm we attribute
to them in a very special case and in a manner unrecognizable as that described
above.  (See \url{https://en.wikipedia.org/wiki/Newton%27s_method}).

For our present situation, the $f(x)$ can be written as a sum of squares as above. It is worth
repeating the vector form of this

$$  f(x) = r^t r $$

because it is less fussy to write and read. This is clear when we expand the evaluation
of the Hessian of this sum of squares function.

@@

then the gradient, in vector form, is

@@

Arguing that the residuals should be "small", Gauss proposed that the 
Newton equations could be approximated by ignoring
the second term (which have elements of the residual vector times second derivatives 
of the residual).  This gives the **Gauss-Newton**
equations (cancelling a factor 2)

@@

Like most other treatments, we will skate over the thin ice that $J^t J$ is a 
good approximation of the Hessian and that the resulting method converges.

@@

with respect to the (scalar) parameter $\alpha$. This one-dimensional minimization can
take quite a lot of computational effort, however, and typically a value of $\alpha$ that
permits a suitable reduction in the sum of squares $f(x)$ is accepted, and the iteration repeated
from

@@

- **A** Compute $J$ and $g$,

- **B** set up Marquardt equations for current $\lambda$ and find $\delta$, 
   compute new set of parameters $$x_{new} = x + \delta $$

@@

Many practitioners dislike using IF statements (comparisons) of floating point
numbers. Tolerances are often suggested to define "small", but this invites users
to make decisions that lead to poor outcomes and which ignore the precision and
other characteristics of the arithmetic at hand.
There are, of course, well-written codes that address the issues of what "small" means,
but such code is generally quite involved. The R function `all.equal()` presents
a well-considered approach. However,
I have never seen a case where the $OFFSET$ approach caused difficulties, and it
is trivial to code. When $x$ is tiny, we compare $OFFSET$ to 
itself and declare equality. When $x$ is very large, $OFFSET$ does not change it and
we compare $x$ with a value that is identical to it.

@@

<!-- ?? Check the statements about how things are done and redo tests -->
Note that we form $J^t J$ in `nlsr::nlfb()` to provide the $D$ matrix for 
the scaled stabilization, 
which is quite a lot of computational work. The same program, however, 
uses a QR approach to solving the linearized least squares sub-problem.
To avoid the formation of $D$, an experimental package
`nlsralt` was been built to add functions `nlxbx()` and `nlfbx()` that
use ONLY the fixed element (i.e., unit matrix) stabilization. A
preliminary timing (using @microbench13) with the Hobbs unscaled problem 
(@cnm79, Example 12.2) shows
a slight advantage for `nlsralt` in execution timing with an equivalent solution.

@@

While the above ideas are the dominant theme in nonlinear least squares, there have been many 
other proposed approaches. The following is a non-exhaustive sample.

@@

@Jones70Spiral seems only to be mentioned in a single paper. There is a flowchart
given but I have never been able to find a code (supposedly Fortran V for the Univac
1108.) The motivations are to avoid both matrix equation solutions and one-dimensional
line-search minimization. Jones mentions matrix inversion, but it is not clear whether
he truly forms the inverse. The main idea in the paper (and presumably the code) is
to use a geometric spiral that mimics the Marquadt search trajectory as $\lambda$
is increased. It would be interesting to actually try some of the ideas, which do,
as far as I can determine, have a reasonable basis and could possibly offer some
efficiencies.

@@

The main attraction of the Gauss-Newton method is that the Jacobian only needs first
derivatives, which are generally much, much easier to compute than the full second
derivatives of the Hessian. However, there have been suggested
Newton-like approaches, possibly for special types of models where some short cut
to the Hessian is available. I regard these approaches as part of the
computational infrastructure of the special cases, as it is usually 
difficult to apply them outside of their context.

Two such examples are the R packages `maxLik` (@maxLik2011) and `trust` (@p-trust).

@@

The central idea of Hartley's method is a line search and that of Marquardt a
parametrized stabilization that can be shown to reduce a sum of squares for a large
enough stabilization parameter. There is nothing to prevent use of a line
search additional to the Marquardt stabilization, and there are, of course, 
several reasonable choices for line search. Also we could solve the (possibly
stabilized) Gauss-Newton equations by several quite different numerical linear
algebra techniques, or even approximations. In certain situations, some of these
choices may be important. Unfortunately, the effort to benchmark the possibilities
seems higher than any benefit in replacing quite good existing methods.

@@

The sources of variety in implementation of nonlinear least-squares methods include:

@@

- sequential or full creation of the Jacobian and residual, since
  this may be done in parts
